Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2025-02-20 paper.tex
Add section on wave action limits

\begin{equation}
\label{eq:phi}
\phi = \dfrac{a_L \omega_L}{k_L} e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
where $z$ is the vertical distance from the mean water level that is positive
upwards.
(\ref{eq:phi}) is accurate to the third order in $\varepsilon_L$ in deep water
because the second- and third-order terms in the Stokes expansion series are
exactly zero.
The long-wave induced horizontal and vertical orbital velocities are:

@@

\begin{equation}
\label{eq:W_L}
W =
\frac{\partial \phi}{\partial z} =
a_L \omega_L e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
Evaluating these velocities at the wave surface $z = \eta_L$ rather than at
the mean water level $z=0$ is a departure from the linear theory, making the
treatment of the long waves here as weakly nonlinear.
As \citet{zhang1990evolution} point out, this is appropriate to do when the
amplitude of the long waves is of the similar magnitude or larger than the
wavelength of the short waves.

@@

@@

\backsection[Acknowledgements]{
I am thankful to Nathan Laxague, Fabrice Ardhuin, Nick Pizzo, and Peisen Tan
for the discussions that helped improve this paper.
I also thank three anonymous reviewers for their helpful comments.
}

@@

\backsection[Data availability statement]{
  The code to run the numerical simulations and generate the figures in this
  paper is available at \url{https://github.com/wavesgroup/wave-modulation-paper}.
  The numerical model is available as a standalone Python package at
  \url{https://github.com/wavesgroup/2wave}.
  %See JFM's \href{https://www.cambridge.org/core/journals/journal-of-fluid-mechanics/information/journal-policies/research-transparency}{research transparency policy} for more information
}

@@

\appendix
\section{Wave action conservation requirements}
\label{appendix:wave_action_conservation}

Following the variational approach by \citet{whitham1965general},
\citet{bretherton1968wavetrains} showed that the wave action is conserved for
small amplitude (linear) waves propagating in slowly-varying media.
Although \citet{longuet1987propagation} correctly pointed out both requirements
(that the medium is slowly-varying and that the short-wave properties are
approximately homogeneous and stationary) for the wave action conservation to
hold, he incorrectly stated that there is no explicit restriction on the
steepness of the long wave, and that it appears necessary to only assume
small-amplitude short waves and significant separation between $k$ and $k_L$.
We will show that in the general case, both the homogeneity and the stationarity
of a short wave quantity $q$ are dependent on the long-wave steepness
$\varepsilon_L$.

\begin{equation}
  \sigma_L \ll \widetilde{\sigma}
\end{equation}
\citet{bretherton1968wavetrains} also require that the medium is nearly uniform
in the vertical, over the vertical scales of motion of the short waves.
For the ambient velocity being the orbital velocity of the long wave, this is
trivially satisfied with (\ref{eq:k_scale_separation}).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_homogeneity_stationarity_analytical.pdf}
\caption{
Homogeneity (left) and stationarity (right) of the short-wave wavenumber (top)
and effective gravity (bottom) as a function of the long-wave steepness
$\varepsilon_L$ and the wavenumber ratio $k/k_L$, based on the linearised
solutions (\ref{eq:homogeneity_k}-\ref{eq:stationarity_g}).
}
\label{fig:homogeneity_stationarity_analytical}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_relative_velocity_error.pdf}
\caption{
}
\label{fig:relative_velocity_error}
\end{figure}

>>>>

\begin{equation}
\label{eq:phi}
\phi = \dfrac{a_L \omega_L}{k_L} e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
where $z$ is the vertical distance from the mean water level that is positive
upwards.
When evaluated at the mean water level ($z=0$), (\ref{eq:phi}) is accurate to
the third order in $\varepsilon_L$ in deep water because the second- and
third-order terms in the Stokes expansion series are exactly zero.
The long-wave induced horizontal and vertical orbital velocities are:

@@

\begin{equation}
\label{eq:W_L}
W =
\frac{\partial \phi}{\partial z} =
a_L \omega_L e^{k_L z} \sin{\psi} + \mathcal{O}(\varepsilon_L^4)
\end{equation}
Evaluating these velocities at the wave surface $z = \eta_L$ rather than at
the mean water level $z=0$ is a departure from the linear theory, making the
treatment of the long waves here as weakly nonlinear.
As \citet{zhang1990evolution} point out, this is appropriate to do when the
amplitude of the long waves is of the similar magnitude or larger than the
wavelength of the short waves.
However, evaluating these velocities at the free surface $z = \eta_L$ introduces
an additional error of $\mathcal{O}(\varepsilon_L^2)$.
The surface velocity errors are described in more detail in Appendix
\ref{appendix:surface_velocity_errors}.

@@

\subsection{Limits to the wave action conservation}
\label{appendix:wave_action_conservation}

Following the variational approach by \citet{whitham1965general},
\citet{bretherton1968wavetrains} showed that the wave action is conserved for
small-amplitude (linear) waves that propagate in slowly-varying media.
\citet{longuet1987propagation} correctly pointed out both
requirements--that both the medium and the short wavetrain must be
slowly-varying--for the wave action conservation to hold.
However, he incorrectly stated that there is no explicit restriction on the
steepness of the long wave, and that it appears necessary to only assume
small-amplitude short waves and significant scale separation ($k/k_L$).
The restriction on $\varepsilon_L$ becomes readily apparent when we recognize
that the long wave-induced orbital velocity scales with $\varepsilon_L$, and
that the divergence of this velocity is the dominant term in (\ref{eq:wavenumber})
and (\ref{eq:wave_action2}).
Here we show that in the general case both the homogeneity and the stationarity
of a short wave quantity $q$ depend on the long-wave steepness $\varepsilon_L$.

\begin{equation}
  \sigma_L \ll \widetilde{\sigma}
\end{equation}
\citet{bretherton1968wavetrains} also require that the medium is nearly uniform
over the vertical scales of motion of the short waves.
For the ambient velocity being the orbital velocity of the long wave, this is
trivially satisfied with (\ref{eq:k_scale_separation}).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_homogeneity_stationarity_analytical.pdf}
\caption{
Homogeneity (left) and stationarity (right) of the short-wave wavenumber (top)
and effective gravity (bottom) as a function of the long-wave steepness
$\varepsilon_L$ and the wavenumber ratio $k/k_L$, based on the linearised
solutions (\ref{eq:homogeneity_k}-\ref{eq:stationarity_g}).
Solid and dashed lines highlight the 0.99 and 0.90 values, respectively.
}
\label{fig:homogeneity_stationarity_analytical}
\end{figure}

Fig. \ref{fig:homogeneity_stationarity_analytical} shows the homogeneities and
stationarities of the short-wave wavenumber and effective gravity based on
the steady solutions and their derived criteria as minima of
(\ref{eq:homogeneity_condition})-(\ref{eq:stationarity_condition}).
Homogeneity is a weaker requirement than stationarity.
At the lowest scale separation considered ($k/k_L = 10$), short-wave wavenumber,
action, and effective gravity are all strongly homogeneous ($H_{k,N,g} > 0.99$).
However, the solutions are only strongly stationary ($S_{k,N,g} > 0.99$) at a
high scale separation of $k/k_L = 100$ for $\varepsilon_L < 0.1$.
If we consider weakly stationary conditions as $S_{k,N,g} > 0.9$, then the
solutions satisfy the requirements for the wave action conservation for
$\varepsilon_L < 0.3$ at $k/k_L = 10$, and for $\varepsilon_L < 0.4$ at
$k/k_L > 20$.
Although \citet{bretherton1968wavetrains} state that the wave action balance
is valid to the first order in $k_L/k$, applying the general criteria
(\ref{eq:homogeneity_condition})-(\ref{eq:stationarity_condition}) yields an
error of $\mathcal{O}(\varepsilon_L k/k_L)$.
We will revisit these criteria based on the numerical solutions of nonlinear
wave action and crest conservation equations in the next section.

@@

\backsection[Acknowledgements]{
I am thankful to Nathan Laxague, Fabrice Ardhuin, Nick Pizzo, and Peisen Tan
for fruitful discussions.
I also thank three anonymous reviewers for their expertise that helped improve
this paper.
}

@@

\backsection[Data availability statement]{
  The code to run the numerical simulations and generate the figures in this
  paper is available at \url{https://github.com/wavesgroup/wave-modulation-paper}.
  The numerical model is available as a standalone Python package at
  \url{https://github.com/wavesgroup/2wave}.
  A Python implementation of SSGW by \citet{clamond2018accurate} is available
  at \url{https://github.com/wavesgroup/ssgw}.
  %See JFM's \href{https://www.cambridge.org/core/journals/journal-of-fluid-mechanics/information/journal-policies/research-transparency}{research transparency policy} for more information
}

@@

\appendix

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_velocity_error_by_ak.pdf}
\caption{
}
\label{fig:velocity_error_by_ak}
\end{figure}
